Phylogenetics

### 2023-05-12 episodes/phylogenetics.Rmd
Complete phylogenetics page.

```{r}
alview(aln)
aln_nogap <- del.colgapsonly(aln, threshold=0.1)
aln_filtered <- aln[,c(7,8,10,12,14,16)]
alview(aln_filtered)
```
Now we have a simple alignment as in the lecture. Dots (`.`) mean that the
sequence is identical to the top row, which makes it easier to calculate 
distances.

@@

Let's now cluster together A and B, and calculate the average distance from
AB to the other sequences.

@@

$d(ABC,D) = \dfrac{d(AB,D) + d(C,D)}{2} = \dfrac{3.5 + 5}{2} = 4.25$

@@

|    | ABC   | D  |
| -  | -:    | -: |
| ABC|       |    |
| D  | 4.25  |    |

@@

The tree is reconstructed by dividing the distances equally between the two
leaves. 
- A-B: each 0.5.
- AB-C: each side gets 2.5/2 = 1.25. The branch to AB is 1.25 - 0.75 = 0.75
- ABC-D: each side gets 4.25/2 = 2.125. The branch to ABC is 2.125 - 0.75 - 0.5 = 0.825

@@

Let's know do the same

@@

```{r}
dist_matrix <- dist.dna(aln_filtered, model="raw") * 6

@@

## Figures

You can include figures generated from R Markdown:

```{r pyramid, fig.alt = "pie chart illusion of a pyramid", fig.cap = "Sun arise each and every morning"}
pie(
  c(Sky = 78, "Sunny side of pyramid" = 17, "Shady side of pyramid" = 5), 
  init.angle = 315, 
  col = c("deepskyblue", "yellow", "yellow3"), 
  border = FALSE
)
```

One of our episodes contains $\LaTeX$ equations when describing how to create
dynamic reports with {knitr}, so we now use mathjax to describe this:

`$\alpha = \dfrac{1}{(1 - \beta)^2}$` becomes: $\alpha = \dfrac{1}{(1 - \beta)^2}$

Cool, right?

>>>>

```{r}
alview(aln)
aln_filtered <- aln[,c(7,8,10,12,14,16)]
alview(aln_filtered)
```
Now we have a simple alignment as in the lecture. Dots (`.`) mean that the
sequence is identical to the top row, which makes it easier to calculate 
distances.

@@

Let's now cluster together A and B, and calculate the average distance 
from AB to the other sequences, weighted by the size of the clusters.

@@

$d(ABC,D) = \dfrac{d(AB,D) * 2 + d(C,D) * 1}{2 + 1} = \dfrac{3.5 * 2 + 5 * 1}{3} = 4$

@@

|    | ABC | D  |
| -  | -:  | -: |
| ABC|     |    |
| D  | 4   |    |

@@

The tree is reconstructed by dividing the distances equally between the two
leaves. 
- A-B: each 0.5.
- AB-C: each side gets 2.5/2 = 1.25. The branch to AB is 1.25 - 0.75 = 0.75
- ABC-D: each side gets 4/2 = 2. The branch to ABC is 2 - 0.75 - 0.5 = 0.75

@@

Let's know do the same using bioinformatics tools.

@@

We'll use `dist.dna` to 
calculate the distances. We'll use a "N" model, that just counts the differences
and doesn't correct or normalizes. We'll use the function `hclust` to perform
the UPGMA method calculation. The tree is then plotted, and the branch 
lengths plotted with `edgelabels`:

```{r}
dist_matrix <- dist.dna(aln_filtered, model="N")
dist_matrix
tree <- as.phylo(hclust(dist_matrix, "average"))
plot(tree)
edgelabels(tree$edge.length)

@@
